A Survey of Quantitative Descriptions of Molecular Structure

### 2012-05-14 paper.tex
More bits

@@

A variety of descriptor implementations are available, with most
cheminformatics toolkits providing implementations of a common set of
descriptors and fingerprints. Table \ref{tab:impl} summarizes the
descriptor classes implemented by several toolkits and
applications. Note that that due to space limitations this is not a
comprehensive list of implemented descriptors across different
applications and the reader is referred to the documentation
accompanying the software packages.

While library calls for descriptor evaluation is the most flexible
approach, it is not useful for non-programmers as well as users who
simply need a descriptor matrix for modeling or searching. As a
result, GUI or command line wrappers around descriptor APIs enable
easier usage. The CDKDescUI application
(\url{http://rguha.net/code/java/cdkdesc.html}) is a Java Swing
application that exposes the CDK descriptor and fingerprint API's. It
allows the users to specify a SMILES or SD file and evaluate a
selected set of descriptors or fingerprints and export them to a text
file. The application also has a command line mode that can be run in
server environments. A similar application is PaDEL \cite{Yap:2011fk}
that includes the CDK descriptors and also implements a number of
additional descriptors such as the extended topochemical atom
descriptors \cite{Roy:2004uq}, McGowan Volume \cite{Zhao:2003kx}, and
the Klekota \& Roth substructure counts \cite{Klekota:2008vn}.

In addition to libraries and their wrappers there are various GUI
applications that allow easy descriptor calculations. Examples include
Bioclipse \cite{Spjuth:2007aa}, MOE (Chemical Computing Group) and
Maestro (Schr\"{o}dinger, Inc.). Workflow tools are also a useful
class of applications for descriptor generation and allow one to
easily generate descriptors from multiple sources. Given the focus of
this article on Open Source implementations, we have noted the
availability in Table \ref{tab:impl}. While a number of
implementations are not strictly Open Source according to OSI
definitions, the fact that they provide free academic licenses does
allow them to be used somewhat freely.

Due to our experience with the CDK project we provide a brief overview
of the architecture of the descriptor calculations implemented by the
CDK. First, the toolkit explicitly differentiates between atom, bond
and whole molecule descriptors. There are a total of 30, 6 and 51
descriptors for atoms, bonds and molecules respectively. Note that
each individual descriptor may calculate multiple values. For example,
the CPSA descriptor evaluates 30 individual values. Thus for a given
molecule it is possible to generate approximately 300 descriptor
values. Every descriptor implements an interface allowing them to
return values in a uniform manner, viz., via a \texttt{DescriptorValue}
object (see Figure~\ref{fig:cdkapi}). This class can represents the result
of a descriptor calculation and includes both the descriptor values
(via a \texttt{IDescriptorResult} object) as well as meta-data on the
descriptor itself via a \texttt{DescriptorSpecification} object (see
below for more details). The descriptor classes are designed such that
in the event of a calculation error (invalid, missing or nonsensical
input), null values are returned, rather than throwing an
exception. This is useful during automated descriptor calculation,
where one desires a ``rectangular'' descriptor matrix, from which one
could filter out the undefined descriptors. While the user can pick
and choose specific descriptors to calculate, the API provides a class
(\texttt{DescriptorEngine}) to specifically evaluate all or a subset
of descriptors for a molecule at one go.  Currently the descriptor
implementations are not threadsafe, making them unsuitable for
parallelization. This is a drawback since many of the descriptors can
be slow. For example, the $\chi$ descriptors \cite{Kier:1976aa} are
implemented using SMARTS matching to identify the various paths and
chains described by Kier \textit{et al.}, rather than explicitly walking the
molecular graph.

@@

As noted above, different implementations of the same descriptor can
generate slightly different values depending on the underlying
chemistry model, differences in reference values and so on. Even
within the same software implementation, different versions of the
same descriptor can generate different values. Thus to maintain
reproducibility in sudies that employ molecular descriptors, some form
of versioning is crucial. The simplest form of version is to manually
note the version of the implementation used for the calculation and
report it along with the results. Though useful, it is not amenable to
automation. In some cases, toolkits will provide an API call to
retrieve the version information, which can then be associated with a
descriptor calculation.

A topic closely related to versioning is descriptor naming. The
preceding sections have discussed a categorization of descriptors and
have used a number of commonly used names. When performing
calculations in an automated manner, or employing remote services for
descriptor calculations \cite{Dong:2007aa}, it is useful to be able to
refer to descriptors via a standardized naming scheme. Currently the
QSAR modeling community lacks such a scheme, though the Chemistry
Development Kit (CDK), a Java library for cheminformatics
\cite{Steinbeck:2006aa} does implement such a scheme. The ``descriptor
specification'' approach defines a set of classes that include
information of the implementation title, identifier and vendor. The
identifier may or may not include a version number. The actual details
of the a descriptor are listed in a dictionary, along with an optional
namespace. The dictionary entry is referenced by the descriptor
specification and this design thus allows one to seamlessly work with
multiple implementations of the same descriptor from different vendors
and keep track of which version is employed throughout a calculation
or workflow. In addition, the CDK provides a hierarchical annotation
of descriptors, grouping them into different classes. Since this
annotation is based on RDF \cite{Taylor:2006ab}, it allows one to go
beyond simple groupings and perform reasoning on descriptors. A simple
example is to identify functionally equivalent descriptors (such as
simple transformations of a base descriptor - square and cube roots of
the gravitational index is an example \cite{Wessel:1998ve}) in an
automated fashion. While this infrastructure is currently only
supported by the CDK, adoption of it (or some similar framework) would
enable easier interoperability between descriptor implementations and
enhance reproducibility.

>>>>

\textbf{Atom and bond descriptors} It should also be noted that various
toolkits also calculate descriptors for chemical entities like atoms and
bonds. These descriptors include simple element property descriptors,
but also geometrical descriptor, such as radial distribution functions
that describe the chemical environment of those atoms. These,
for example, have been used to predict proton shift in 
nuclear magnetic resonance~\cite{AiresDeSousa2002}. Similarly, bond
descriptors can be used to predict bond properties.

@@

A variety of descriptor implementations are available, with most
cheminformatics toolkits providing implementations of a common set of
descriptors and fingerprints. Table~\ref{tab:impl} summarizes the
descriptor classes implemented by several toolkits and
applications. Note that due to space limitations this is not a
comprehensive list of implemented descriptors across different
applications and the reader is referred to the documentation
accompanying the software packages.

While library calls for descriptor evaluation is the most flexible
approach, it is not useful for non-programmers as well as users who
simply need a descriptor matrix for modeling or searching. As a
result, GUI or command line wrappers around descriptor APIs enable
easier usage. The CDKDescUI application
(\url{http://rguha.net/code/java/cdkdesc.html}) is a Java Swing
application that exposes the Chemistry Development Kit (CDK) descriptor and fingerprint API's. It
allows the users to specify a SMILES or SD file and evaluate a
selected set of descriptors or fingerprints and export them to a text
file. The application also has a command line mode that can be run in
server environments. A similar application is PaDEL \cite{Yap:2011fk}
that includes the CDK descriptors and also implements a number of
additional descriptors such as the extended topochemical atom
descriptors \cite{Roy:2004uq}, McGowan Volume \cite{Zhao:2003kx}, and
the Klekota \& Roth substructure counts \cite{Klekota:2008vn}.

In addition to libraries and their wrappers there are various GUI
applications that allow easy descriptor calculations. Examples include
Bioclipse \cite{Spjuth:2007aa}, MOE (Chemical Computing Group) and
Maestro (Schr\"{o}dinger, Inc.). Workflow tools are also a useful
class of applications for descriptor generation and allow one to
easily generate descriptors from multiple sources~\cite{Truszkowski2011}. Given the focus of
this article on Open Source implementations, we have noted the
availability in Table \ref{tab:impl}. While a number of
implementations are not strictly Open Source according to OSI
definitions, the fact that they provide free academic licenses does
allow them to be used somewhat freely.

Due to our experience with the CDK project we provide a brief overview
of the architecture of the descriptor calculations implemented by the
CDK. First, the toolkit explicitly differentiates between atom, bond
and whole molecule descriptors. There are a total of 30, 6 and 51
descriptor algorithms for atoms, bonds and molecules respectively. Note that
each individual descriptor algorithm may calculate multiple descriptors. For example,
the CPSA descriptor evaluates 30 individual descriptor values. Thus for a given
molecule it is possible to generate approximately 300 descriptor
values.

Every descriptor implements an interface allowing them to
return values in a uniform manner, viz., via a \texttt{DescriptorValue}
object (see Figure~\ref{fig:cdkapi}). This class can represents the result
of a descriptor calculation and includes both the descriptor values
(via a \texttt{IDescriptorResult} object) as well as meta-data on the
descriptor itself via a \texttt{DescriptorSpecification} object (see
below for more details). The descriptor classes are designed such that
in the event of a calculation error (invalid, missing or nonsensical
input), null values are returned, rather than throwing an
exception. This is useful during automated descriptor calculation,
where one desires a ``rectangular'' descriptor matrix, from which one
could filter out the undefined descriptors. While the user can pick
and choose specific descriptors to calculate, the API provides a class
(\texttt{DescriptorEngine}) to specifically evaluate all or a subset
of descriptors for a molecule at one go.  Currently the descriptor
implementations are not threadsafe, making them unsuitable for
parallelization. This is a drawback since many of the descriptors can
be slow. For example, the $\chi$ descriptors \cite{Kier:1976aa} are
implemented using SMARTS matching to identify the various paths and
chains described by Kier \textit{et al.}, rather than explicitly walking the
molecular graph.

@@

As noted above, different implementations of the same descriptor can
generate slightly different values depending on the underlying
chemistry model, differences in reference values and so on. Even
within the same software implementation, different versions of the
same descriptor can generate different values. Thus to maintain
reproducibility in studies that employ molecular descriptors, some form
of versioning is crucial. The simplest form of version is to manually
note the version of the implementation used for the calculation and
report it along with the results. Though useful, it is not amenable to
automation. In some cases, toolkits will provide an API call to
retrieve the version information, which can then be associated with a
descriptor calculation.

A topic closely related to versioning is descriptor naming. The
preceding sections have discussed a categorization of descriptors and
have used a number of commonly used names. When performing
calculations in an automated manner, or employing remote services for
descriptor calculations \cite{Dong:2007aa}, it is useful to be able to
refer to descriptors via a standardized naming scheme. Currently the
QSAR modeling community lacks such a scheme, though the CDK, a Java library for cheminformatics
\cite{Steinbeck:2006aa} does implement such a scheme. The ``descriptor
specification'' approach defines a set of classes that include
information of the implementation title, identifier and vendor. The
identifier may or may not include a version number. The actual details
of the a descriptor are listed in a dictionary, along with an optional
namespace. The dictionary entry is referenced by the descriptor
specification and this design thus allows one to seamlessly work with
multiple implementations of the same descriptor from different vendors
and keep track of which version is employed throughout a calculation
or workflow. In addition, the CDK provides a hierarchical annotation
of descriptors, grouping them into different classes. Since this
annotation is based on RDF \cite{Taylor:2006ab}, it allows one to go
beyond simple groupings and perform reasoning on descriptors. A simple
example is to identify functionally equivalent descriptors (such as
simple transformations of a base descriptor - square and cube roots of
the gravitational index is an example \cite{Wessel:1998ve}) in an
automated fashion. While this infrastructure is currently only
supported by the CDK, adoption of it (or some similar framework) would
enable easier interoperability between descriptor implementations and
enhance reproducibility. The framework is, however, actively used by
software using the CDK descriptors, such as
Bioclipse~\cite{Spjuth2010} and Ambit~\cite{Jelizakova2011}. Moreover,
the ideas have been adopted by the CHEMINF ontology for
cheminformatics~\cite{Hastings2011}, which is used by the CHESS
framework which also uses the CDK for descriptor calculations~\cite{Chepelev2011b}.
